---
title: "PML Prediction Assignment"
author: "mandytcs"
date: "27 Feb 2016"
output: html_document
---

###Background: 
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, my goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

###Data:
The training data for this project are available here:
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here:
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data for this project come from this source: http://groupware.les.inf.puc-rio.br/har. 

###Prepare the datasets:
The following packages are required for this analysis.
```{r message=FALSE, warning=FALSE}
library(caret)
library(dplyr)
library(randomForest)
```
Note: assuming that the test and training csv data sets are contained in the working directory.
Read the training data into R, identifying "NA", "" and "#DIV/0!" as NA strings.
```{r}
pmltrain <- read.csv("pml-training.csv",na.strings=c("NA","","#DIV/0!"))
```

###Spliting the data into training and test sets:
Split the plmtrain into a training set (for model training) and a test set (for predicting the out of sample error), splitting on the classe variable (this is the variable of interest) with a 70-30 split
```{r}
set.seed(555)
# Taking 70% for the training data and 30% for the test data
inTrain <- createDataPartition(y = pmltrain$classe, list = FALSE, p=0.7)
trainData <- pmltrain[inTrain,]
testData <- pmltrain[-inTrain,]
```
The analysis is now conducted purely on the trainData until the model is build and an out of sample error is needed.

Identify variables that are mostly NAs:
```{r}
table(is.na(trainData))
```
Find which variables (if any) that are mostly na values
```{r}
naprops <- colSums(is.na(trainData))/nrow(trainData)
mostlyNAs <- names(naprops[naprops > 0.75]) # mostly being 75%
mostlyNACols <- which(naprops > 0.75) # there's about 100 of them
```
Take a small sample of the training data to work with
```{r}
set.seed(1256)
smalltrain <- trainData %>% tbl_df %>% sample_n(size=1000)
```
Remove the variables that are made up of mostly NAs
```{r}
smalltrain <- smalltrain[,-mostlyNACols]
```
Remove the row number (X) and user_name column
```{r}
smalltrain <- smalltrain[,-grep("X|user_name",names(smalltrain))]
```
This factor variable makes prediction of the test set difficult and is reduandant when raw time data is available in the data set.
```{r}
smalltrain <- smalltrain[,-grep("cvtd_timestamp",names(smalltrain))]
```
Remove candidate predictors that have near zero variance
```{r}
smalltrain <- smalltrain[,-nearZeroVar(smalltrain)]
```
###List of candidate predictors:
```{r}
modelVars <- names(smalltrain)
modelVars1 <- modelVars[-grep("classe",modelVars)] # remove the classe var
```
The predictors for the machine learning are:
```{r}
modelVars1
```

###Build a random forest model:
Using a random forest with the predictors in modelVars1 to predict the classe variable.
```{r}
set.seed(57)
cleanedTrainData <- trainData[,modelVars]
modelFit <- randomForest(classe ~., data=cleanedTrainData, type="class")
```
Get Error Estimates:
Begin with an insample error estimate (from trainData - which is 70% of pml-training.csv)
```{r}
## Get the values predicted by the model
predTrain <- predict(modelFit,newdata=trainData)

## Use a confusion matrix to get the insample error
confusionMatrix(predTrain,trainData$classe)$table
```
The in sample error is unrealistically high.
Now getting an out of sample error estimate (from testData - which is 30% of pml-training.csv)
```{r}
classe_col <- grep("classe",names(testData))
predTest <- predict(modelFit, newdata = testData[,-classe_col], type="class")

confusionMatrix(predTest,testData$classe)
```
The model has an out of sample accuracy of: 0.998.

###Prediciting exercise activity using the model
Load the pml-test data
```{r}
pmltest <- read.csv("pml-testing.csv",na.strings=c("NA","","#DIV/0!"))
```
Perform the prediction
```{r}
# plmtest predicition
predplmtest <- predict(modelFit, newdata = pmltest, type="class")
print(predplmtest)
```
###Conclusion
Random Forest return a high accuraccy of 99% to predict proper weight lifting fitting accelerometer data.
Random Forest fitting predicts correctly every observation over the 20 test cases.